\emph{icepack}: a novel glacier flow modeling package

### 2019-10-06 icepack.tex
More on convex opt

One of our goals with icepack is to facilitate experimentation with the model physics.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
In C++ one could avoid modifying the diagnostic solve routine by using variadic templates.
Variadic templates are a more advanced language feature and relying on them would present a steep learning curve for novices.

@@

The action functional for each diagnostic model is convex, i.e. the second derivative is positive-definite.
\textcolor{red}{Why it's great...}

@@

Calculating the ice velocity from the geometry and other input fields involves minimizing a convex functional.
The minimizer can only be approximated, so we employ iterative procedures based on Newton's method with a line search.
We must then choose some metric to decide when the current guess for the velocity is good enough to stop the iteration.

A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
The search direction $v_k$ at step $k$ for Newton's method is defined as
\begin{equation}
    v_k = -\ud^2J(u_k)^{-1}\ud J(u_k).
\end{equation}
Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, this is a descent direction for the action $J$:
\begin{equation}
    \ud J(u_k)\cdot v_k < 0.
\end{equation}
The absolute value of the quantity in the last equation is defined as the Newton decrement.
For $u_k$ sufficiently close to the true solution $u$, the Newton decrement roughly tells us how much we can expect the action to decrease:
\begin{equation}
    J(u_k) - J(u) \approx \frac{1}{2}|\ud J(u_k)\cdot v_k|.
\end{equation}
We can then use the Newton decrement to decide when to stop the iteration.

>>>>

One of our goals with icepack is to facilitate experimentation with the model physics.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms, this amounts to changing the number of arguments to the function that calculates basal shear stress.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
In C++ one could avoid modifying the diagnostic solve routine by using variadic templates, but this is an advanced language feature and relying on it would present a steep learning curve for novices.

@@

The action functional for each diagnostic model is convex, i.e. the second derivative is strictly positive-definite.
From a theoretical persepctive, convexity guarantees that the problem has a unique solution.
This property is also especially advantageous for implementing numerical solvers.
We use a damped Newton method to solve the diagnostic equations.
Starting from a guess $u_k$ for the velocity, one first computed a \emph{search direction} $v_k$ that solves a linear equation:
\begin{equation}
    \ud^2J(u_k)\cdot v_k = -\ud J(u_k).
    \label{eq:newton-search-direction}
\end{equation}
The next value of the velocity is then computed by searching along $v_k$:
\begin{equation}
    u_{k + 1} = u_k + \alpha_k\cdot v_k,
\end{equation}
where $\alpha$ is chosen to minimize $J$ along the search direction.
For an initial guess sufficiently close to the exact solution, the undamped Newton method ($\alpha_k = 1$ at every step) converges quadratically.
The line search step ensures that the method can converge even from a poor initial guess, provided that the line search method satisfies the Armijo-Wolfe criteria \citep{nocedal2006numerical}.

For a convex problem, $\ud^2J$ is a symmetric and positive-definite matrix.
This has two advantages.
First, the search direction is always a descent direction for $J$.
For non-convex problems this is not always the case and some adaptation strategy must be employed in the vicinity of saddle points.
Second, symmetry and positivity enable the use of specialized linear solvers, such as the Cholesky decomposition or the conjugate gradient algorithm, that are superior to their more general counterparts in many respects.

@@

Several works in the literature have weighed the relative merits of different iterative methods for solving the nonlinear diagnostic equation, but few if any have considered the problem of when to stop iterating.
The most common stopping criteria are when (1) the 2-norm of the residual is sufficiently small or (2) the relative change in the iterates is sufficiently small.
Each of these approaches has problems.
The residual norm depends on the discretization and does not weight all degrees of freedom proportionally, e.g. vertex and edge degrees of freedom in higher-order finite element methods.
The relative change criterion, on the other hand, can suggest convergence when in fact the method has stagnated.

A convergence criterion that works equally well independent of the mesh, finite element discretization, and the quality of the initial guess can be defined based on the idea of the \emph{Newton decrement} \citep{nocedal2006numerical}.
Since the second derivative operator $\ud^2J(u_k)$ is positive-definite, the Newton search direction $v_k$ computed from equation \eqref{eq:newton-search-direction} is a descent direction for $J$:
\begin{equation}
    \ud J(u_k)\cdot v_k < 0.
\end{equation}
The absolute value of the quantity in the last equation is defined as the Newton decrement.
For $u_k$ sufficiently close to the true solution $u$, the Newton decrement roughly tells us how much we can expect the action to decrease:
\begin{equation}
    J(u_k) - J(u) \approx \frac{1}{2}|\ud J(u_k)\cdot v_k|.
\end{equation}
We can then use the Newton decrement to decide when to stop the iteration.
